---
title: The occurence of Clemmys guttata (Spotted Turtle) with respect to wetland habitat
  suitability
author: "Matthew Amoia"
subtitle: Presence/Absence of spotted turtle with wetlands dynamics
---

# Introduction
Habitat fragmentation and degradation are undoubtedly a critcial player in the population dynamics of many species, due to humans manipulation of natural ecosystems. As the land cover on Earth is altered so drastically, we can witness species trying to react to such changes. This results in understanding how these species may or may not be able to react, resulting in conservation efferts. Understanding and assessing the dynamics of changes in wetlands and land cover provide evidence for species occurrence and where organisms will have suitable habitat to live. The spotted turtle is a small sized turtle with black coloration on the carapace, and bright yellow spots (Gibbs et al. 2007). They inhabit vernal pools in the spring, and upland forests in the summer, also inhabiting wet meadows, forested swamps, and sphagnum bogs (Gibbs et al. 2007). The spotted turtle can be found on the Great Lakes Plains south of Lake Ontario and east of Lake Erie, the lower Hudson river valley, and on Long Island. Aside from New York, spotted turtles are found as far north as Quebec, Canada, down to the state of Florida, and over to the central plains such as Illionis (Gibbs et al. 2007). The spotted turtle is declining throughout much of its range, and it is listed as a species of Special Concern in New York State. In New York the spotted turtle suffers from habitat fragmentation, wetland draining, suburban development, invasive wetland plants, road mortality (Lewis et al. 2004). Buchanan et al. 2017 states that their can be positive reactions from spotted turtles when specific parts of their habitat are specifically manipulated, however, Markle et al. 2018 warns that long-term habitat changes within specified habitats have implications for herpetofauna habitat management and restoration. With the assessment of wetlands in concert with occurrence citizen science data, the scope of this study is to address potential problems within habitat suitablity for the spotted turtle.      

# Hypothesis/s
The changes in wetlands habitat will result in less occurence of the spotted turtle. Areas of high land use changes will result in the displacement of spotted turtles, and places where wetlands hang on will possibly have the greatest occurrence of spotted turtles. 

# Materials and methods

Materials- 

Data: Occurence data for the spotted turtle will come from the Global Biodiversity Information Facility (GBIF), and the wetlands data will come from the US Fish and Wildlife Service National Wetlands Inventory. 

Methods: 

* Visualize the occurence of spotted turtles.
* Visualize the presence of wetlands and their dynamic changes.
* Correlate and overlay these occurences to understand and potenially provide insight into where spotted turtles are found, and where they may be found based on the changing dynamics of the nations wetlands. 

Load any required packages (you may need to install some packages):

```{r, loading packages, message=F, warning=F}
library(tidyverse)
library(maps)
library(spocc)
knitr::opts_chunk$set(cache=TRUE)  # cache the results for quick compiling 
```

The US Fish and Wildlife Service National Wetlands Inventory Data will either be downloaded by HUC8 Watershed Boundaries or by State.

Use of `occ()` function to download _occurrence_ observations for the Spotted turtle (_Clemmys guttata_) from the [Global Biodiversity Information Facility](gbif.org).


```{r, warning=F} 
## define which species to query
sp='Clemmys guttata'

## run the query and convert to data.frame()
d = occ(query= 'Clemmys guttata', from = 'gbif', limit = 1000) %>% occ2df(what = "data") 
```

```{r, visualizing occurence data}

ggplot(d,aes(x=longitude,y=latitude))+geom_point() 
  
```

# Results
The results will show the occurence gbif of spotted turtles based on the citizen science program iNaturalist, and will hopefully shed light on where these creatures are present. Furthermore, the US fish and Wildlife Service National Wetlands Inventory will allow for the understanding of how the current projection of wetlands is correlated or not with the current obsrevation projection of the spotted turtle. 

# Conclusions
My take away from this project is to hopefully understand the geographical representation of the spotted turtle. Futhermore, with the addition of the wetlands inventory, I hope to point out the possible "hotspots", and a general overview of the conservation and biogeography of the spotted turtle. 

# References

Buchanan, Scott W., Bill Buffum, Karraker, Nancy E. Responses of a spotted turtle (Clemmys guttata) population to creation of early-sucessional habitat. 16 December 2017. Herpetological Conservation and Biology. 12(3):688-700.

Gibbs, James P., Breisch, Alvin R., Ducey, Peter K., Glenn Johnson, Behler, John L., Bothner, Richard C. 2007. The Amphibians and Reptiles of New York State Identifcation, Natural History, and Conservation. Oxford University Press. Pages 173-176. 

Lewis, T.L., J.M. Ullmer, and J.L. Mazza. 2004. Treats to spotted turtle (Clemmys guttata) habitat in Ohio. The Ohio Journal of Science 104:65-71. 

Markle, Chantel E., Gillian Chow-Fraser, Patricia Chow-Fraser. Long-term habitat changes in a protected area: Implcations for herpetofauna habitat management and restoration. February 14, 2018. Plos One. https://doi.org/10.1371/journal.pone.0192134  
